In-silico molecular enrichment and clearance of the human intracranial space

### 2025-06-05 paper/main.tex
Fix subsection title.
Add thank you.
Move Methods and revamp transport description. WIP.

We model diffusion, convection, and exchange of a molecular solute in
the PVS networks $\Lambda_a$ and $\Lambda_v$, in the CSF spaces
$\Omega_{\rm CSF}$ and in the brain parenchyma $\Omega_{\rm PAR}$ via
a mixed-dimensional transport model~\cite{masri2024modelling,laurino2019derivation} over a
timescale of minutes to hours (\Cref{fig:results1}A--B,
\Cref{fig:csf}A--C). Specifically, for $t > 0$, we solve for a
concentration $c = c(x, t)$ in the 3D intracranial compartments ($x
\in \Omega_{\rm CSF}, x \in \Omega_{\rm PAR}$) and for a cross-section
averaged concentration $\hat{c} = \hat{c}(s, t)$ in the periarterial
and perivenous networks ($s \in \Lambda_a, s \in \Lambda_v$) such that
the following equations hold:
\begin{subequations}
\begin{alignat}{2}
  \partial_t (\phi c) - \nabla \cdot (D \nabla (\phi c) ) + \nabla \cdot (\bm u c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma & = 0 && \quad \quad \mathrm{in} \quad \Omega_{\rm CSF}, \Omega_{\rm{PAR}},
  \label{eq:multi_transport_3d}
  \\ 
  \partial_t (A  \hat c) - \partial_s(\hat D A \partial_s ( \hat c)) +\partial_s(A \hat u \hat c )  +  \xi P (\hat c - \overline{c})  &= 0 && \quad \quad \mathrm{in} \quad  \Lambda_a, \Lambda_v .
  \label{eq:multi_transport_1d}
 \end{alignat}%
\label{eq:multi_transport}%
\end{subequations}%
In~\eqref{eq:multi_transport}, $0 < \phi \leqslant 1$ is the fluid
volume fraction, $D$ and $\hat{D}$ are effective diffusion
coefficients, and $\bm u$ and $\hat{u}$ are fluid (CSF/ISF)
velocities, all defined in the respective compartments; $P = 2 \pi
R_2$ is the perimeter and $A = \pi (R_2^2 - R_1^2)$ the area of the
PVS cross-sections, $\delta_{\Gamma}$ is a Dirac term concentrated on
the interfaces $\Gamma_a, \Gamma_v$, the overlined $\overline{c}$
denotes the average of $c$ over cross-sections of $\Gamma_a,
\Gamma_v$, and $\xi$ is a permeability allowing for transfer/exchange
over the interfaces $\Gamma_a, \Gamma_v$ between the periarterial and
perivenous networks and their surroundings.  Moreover, we model the
membrane between the parenchyma and the CSF spaces as a semi-permeable
membrane with permeability coefficient $\beta_{\rm pia}$. For a
complete description of this mathematical model and simulation
scenarios, see the Supplementary information S1. The initial concentrations
in the three-dimensional domain $\Omega$ and networks $\Lambda_a,
\Lambda_v$ are all set to zero.

@@

@@

\subsection*{Model validation}

@@

All authors express their gratitude to Prof.~Jürgen Reichenbach for
generously sharing the multi-modal human MR data, and to Prof. Erlend
Hodneland for facilitating this data exchange. M.E.R.~acknowledges
support from Stiftelsen Kristian Gerhard Jebsen via the K.~G.~Jebsen
Centre for Brain Fluid Research. M.C., R.M., and M.E.R.~acknowledge
support from the Research Council of Norway (RCN) via grant \#324239
(EMIx). Simulations were performed on the Experimental Infrastructure
for Exploration of Exascale Computing (eX3), which is financially
supported by the RCN under contract 270053. M.K.~acknowledges support
from the RCN grant \#303362. The figures 1A, 1D, 2C, 3C, 5A and 6A
were created in Biorender.

@@

M.C., M.K, R.M., and M.E.R. conceived and designed the project.
M.C., M.K, R.M., and M.E.R. contributed to software development. 
M.C., M.K, and R.M. conducted the experiments.
M.C. analyzed the results and prepared the figures.
M.C. M.K. and R.M and M.E.R wrote the manuscript. 
All authors edited and reviewed the manuscript.

>>>>

@@

We represent the evolution and distribution of a solute in the CSF
spaces $\Omega_{\rm CSF}$, brain parenchyma $\Omega_{\rm PAR}$, and
periarterial and perivenous networks $\Lambda_a \cup \Lambda_v$ via a
mixed-dimensional transport model~\cite{masri2024modelling} over a timescale
of minutes to hours (\Cref{fig:results1}A--B, \Cref{fig:csf}A--C). The
current model accounts for solute exchange between the CSF spaces,
brain parenchyma, and the PVSs, diffusion within all compartments,
dispersion and convection in the CSF spaces, and convection within the
PVS. Specifically, for $t > 0$, we solve for a concentration field
$c(x, t)$ in the 3D intracranial compartments (for $x \in \Omega_{\rm
  CSF} \cup \Omega_{\rm PAR}$) and for a cross section-averaged
concentration field $\hat{c}(s, t)$ in the perivascular networks (for
$s \in \Lambda_a, \Lambda_v$) satisfying the following equations:
\begin{subequations}
\begin{alignat}{2}
  \partial_t c - \nabla \cdot D (1 + R) \nabla c  + \nabla \cdot (\bm u_{\rm CSF} c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma & = 0 && \quad \quad \mathrm{in} \quad \Omega_{\rm CSF},
  \label{eq:multi_transport_3d_CSF}
  \\ 
  \partial_t (\phi c) - \nabla \cdot D^{\ast} \nabla (\phi c)  + \nabla \cdot (\bm u_{\rm PAR} c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma & = 0 && \quad \quad \mathrm{in} \quad \Omega_{\rm{PAR}},
  \label{eq:multi_transport_3d_PAR}
  \\ 
  \partial_t (A  \hat c) - \partial_s(\hat D A \partial_s ( \hat c)) +\partial_s(A \hat u \hat c )  +  \xi P (\hat c - \overline{c})  &= 0 && \quad \quad \mathrm{in} \quad  \Lambda_a, \Lambda_v .
  \label{eq:multi_transport_1d}
 \end{alignat}%
\label{eq:multi_transport}%
\end{subequations}%
In~\eqref{eq:multi_transport_3d_CSF}, $D$ is the (free) diffusion
coefficient of the solute, $R$ is a dispersion coefficient, $\bm
u_{\rm CSF}$ is a CSF velocity, $\xi$ is a permeability allowing for
transfer/exchange across the interfaces $\Gamma_a, \Gamma_v$ between
the periarterial and perivenous networks and their surroundings, the
overlined $\overline{c}$ denotes the average of $c$ over
cross-sections of $\Gamma_a, \Gamma_v$, and $\delta_{\Gamma}$ is a
Dirac term concentrated on $\Gamma_a, \Gamma_v$. Additionally, in
\eqref{eq:multi_transport_3d_PAR}, $\phi$ is the brain extracellular
volume fraction\footnote{We set $\phi = 1.0$ in $\Omega_{\rm CSF}$ and
therefore omit $\phi$ in~\eqref{eq:multi_transport_3d_CSF}.},
$D^{\ast}$ is the effective diffusion coefficient in the parenchyma,
and $\bm u_{\rm PAR}$ is an interstitial fluid velocity. In this
study, we set $\bm u_{\rm PAR} = 0$. In~\eqref{eq:multi_transport_1d},
$A = \pi (R_2^2 - R_1^2)$ is the area and $P = 2 \pi R_2$ is the
perimeter of the PVS cross-sections, $\hat{D}$ is the effective
diffusion coefficient in the PVSs, and $\hat{u}$ is the fluid velocity
(in the axial direction) in the PVS. Further, we model the membrane
between the parenchyma and the CSF spaces as a semi-permeable membrane
with permeability coefficient $\beta_{\rm pia}$. For an extended description of the mathematical model and simulation
scenarios, see the Supplementary information S1. The initial
concentrations in the three-dimensional domain $\Omega$ and networks
$\Lambda_a, \Lambda_v$ are all set to zero.

%% \subsection*{Molecular transport equations}

%% We model diffusion, convection, and exchange of a molecular solute in
%% the PVS networks $\Lambda_a$ and $\Lambda_v$, in the CSF spaces
%% $\Omega_{\rm CSF}$ and in the brain parenchyma $\Omega_{\rm PAR}$ via
%% a mixed-dimensional transport model~\cite{masri2024modelling,laurino2019derivation} over a
%% timescale of minutes to hours (\Cref{fig:results1}A--B,
%% \Cref{fig:csf}A--C). Specifically, for $t > 0$, we solve for a
%% concentration $c = c(x, t)$ in the 3D intracranial compartments ($x
%% \in \Omega_{\rm CSF}, x \in \Omega_{\rm PAR}$) and for a cross-section
%% averaged concentration $\hat{c} = \hat{c}(s, t)$ in the periarterial
%% and perivenous networks ($s \in \Lambda_a, s \in \Lambda_v$) such that
%% the following equations hold:
%% \begin{subequations}
%% \begin{alignat}{2}
%%   \partial_t (\phi c) - \nabla \cdot (D \nabla (\phi c) ) + \nabla \cdot (\bm u c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma & = 0 && \quad \quad \mathrm{in} \quad \Omega_{\rm CSF}, \Omega_{\rm{PAR}},
%%   \label{eq:multi_transport_3d}
%%   \\ 
%%   \partial_t (A  \hat c) - \partial_s(\hat D A \partial_s ( \hat c)) +\partial_s(A \hat u \hat c )  +  \xi P (\hat c - \overline{c})  &= 0 && \quad \quad \mathrm{in} \quad  \Lambda_a, \Lambda_v .
%%   \label{eq:multi_transport_1d}
%%  \end{alignat}%
%% \label{eq:multi_transport}%
%% \end{subequations}%
%% In~\eqref{eq:multi_transport}, $0 < \phi \leqslant 1$ is the fluid
%% volume fraction, $D$ and $\hat{D}$ are effective diffusion
%% coefficients, and $\bm u$ and $\hat{u}$ are fluid (CSF/ISF)
%% velocities, all defined in the respective compartments; $P = 2 \pi
%% R_2$ is the perimeter and $A = \pi (R_2^2 - R_1^2)$ the area of the
%% PVS cross-sections, $\delta_{\Gamma}$ is a Dirac term concentrated on
%% the interfaces $\Gamma_a, \Gamma_v$, the overlined $\overline{c}$
%% denotes the average of $c$ over cross-sections of $\Gamma_a,
%% \Gamma_v$, and $\xi$ is a permeability allowing for transfer/exchange
%% over the interfaces $\Gamma_a, \Gamma_v$ between the periarterial and
%% perivenous networks and their surroundings.  Moreover, we model the
%% membrane between the parenchyma and the CSF spaces as a semi-permeable
%% membrane with permeability coefficient $\beta_{\rm pia}$. For a
%% complete description of this mathematical model and simulation
%% scenarios, see the Supplementary information S1. The initial concentrations
%% in the three-dimensional domain $\Omega$ and networks $\Lambda_a,
%% \Lambda_v$ are all set to zero.

@@

\subsection*{Comparison with literature results}

@@

All authors express their gratitude to Prof.~Jürgen Reichenbach for
generously sharing the multi-modal human MR data, and to Prof. Erlend
Hodneland for facilitating this data exchange. The authors also thank
Profs.~Per Kristian Eide, Geir Ringstad and Rune Enger for valuable
input and discussions. M.E.R.~acknowledges support from Stiftelsen
Kristian Gerhard Jebsen via the K.~G.~Jebsen Centre for Brain Fluid
Research. M.C., R.M., and M.E.R.~acknowledge support from the Research
Council of Norway (RCN) via grant \#324239 (EMIx). Simulations were
performed on the Experimental Infrastructure for Exploration of
Exascale Computing (eX3), which is financially supported by the RCN
under contract 270053. M.K.~acknowledges support from the RCN grant
\#303362. The figures 1A, 1D, 2C, 3C, 5A and 6A were created in
Biorender.

@@

M.C., M.K, R.M., and M.E.R. conceived and designed the project.
M.C., M.K, R.M., and M.E.R. contributed to software development. 
M.C., M.K, and R.M. conducted the experiments.
M.C. analyzed the results and prepared the figures.
M.C., M.K., R.M and M.E.R wrote the manuscript. 
All authors edited and reviewed the manuscript.
